The low dimensional world of echolocating bats

### 2020-03-25 Paper/main.tex
updated paper based on herbert comments

Echolocating bats are long-lived, mobile animals. They use the landscape intensively for foraging, roosting, and commuting. They have excellent spatial memory \citep{Barchi2013,VonHelversen2005} and can rely on sonar to navigate to notable locations like foraging grounds, drinking places \citep[see][for references]{Vanderelst2016,Vanderelst2017}. Furthermore, bats have been reported to use sonar-based recognition of landmarks, both under experimental \citep{Jensen2005,Yu2019}, and natural \citep{Verboom1999} conditions.

@@

Together these results suggest that echolocating bats live in a low dimensional perceptual world. The physics of the sonar system and the auditory periphery map the complexity of the 3D environment onto a small number of independent features, irrespective of whether we use artificial or real echo sequences. Nevertheless, this limited set of features is sufficiently rich for bats to recognize locations, and as we discuss, support flight control.

@@

We created artificial echo sequences by randomly spacing ideal point reflectors. The distances of the point reflectors were selected using a two-step process. First we selected $n_{s}$ seed points $s$, uniformly distributed over a range from 0.5 to 7 meters. Next, for every seed point $s$, we selected $n_{c}$ cloud points normally distributed around $s$, with standard deviation 0.25 m. This two step process to modeling natural reflectors was inspired by the results of \citet{Yovel2009}, who found that echoes returning from vegetation could be better modeled using a similar two-step process than simply using uniformly spaced points. We generated xx sets of points for each combination of $n_s \in [xx,xx,xx, xx]$ and $n_c \in [xx, xx, xx, xx]$. Synthetic echoes consisting of few reflectors are likely to consist of only weak echoes. To ensure that all sequences included at least one strong echo, we added a single reflector at a distance between xx and xx to each collection of points.

Each set of points was converted to an impulse response with a sampling frequency $f_s$ of 219 kHz. The impulse response length was set to 7499 samples. These parameters correspond to the recording length and the sampling frequency used in the acoustic data collection (see below). In converting the point to an impulse response, we modeled attenuation of the echoes due to spherical spreading (both ways). We sampled the amplitude of the impulses from a normal distribution with a standard deviation of 10 dB.

The impulse responses were converted to echo sequences by convolving them with a bat-like emission. We used a 1 millisecond long linear frequency-swept cosine starting at 100 kHz and ending at 40 Khz. Again, these parameters correspond to those used in the acoustic data collection described below. The echo sequences were dechirped as follows, 
%
%ifft(complex conj(ftt emission) .* fft(echo))
%
\begin{equation}\label{key}
E_d \mathcal{F}^{-1}(\mathcal{F}(S)^* . \mathcal{F}(E))
\end{equation}
with $\mathcal{F}$ denoting the Fourier transform and $()^*$ the complex conjugate. The symbols $S$, $E$, $E_d$ refer to the emission, the echo sequence and the dechirped echo sequence, respectively. Examples of the resulting echo sequences are shown in figures \ref{fig:processing} and \ref{fig:synthetictemplates}.

Artificial echo sequences were filtered using a model of the bat's auditory periphery, as proposed by \citet{Wiegrebe2008}. This model simulates the temporal and spectral resolution of the bat's cochlea. The model consists of a gammatone filterbank followed by half-wave rectification and non-linear compression with an exponent of 0.4. Finally, the model includes a 1 kHz low-pass filter modeling the response characteristics of the inner hair cells. In the current implementation, we used a gammatone filterbank with xx filters, centered around xx kHz. The filters were spaced to correspond equivalent rectangular bandwidths \citep{Glasberg1990}. The output of each channel was filtered with a time (distance) weighted factor corresponding to the atmospheric attenuation at the channel's center frequency. This filtering approximated frequency-dependent atmospheric attenuation. The first 6 ms of data were removed. This was done to ensure the data corresponded with the acoustic data. As indicated below, we removed the first 6 ms of those data to avoid including the pickup of the emitted signal.

Figure \ref{fig:processing} depicts a single example of a synthetic echo sequence. It also illustrates the resulting response for each of the xx channels of the \citet{Wiegrebe2008} model. From these responses, it can be seen that these are highly correlated across channels. This is confirmed when analyzing the average cross-channel correlation for all xx synthetic echo sequences (Fig. \ref{fig:processing}xx). From these high cross-channel correlations, we conclude that the output of the cochlear model can be summarized by averaging across channels without extensive loss of information. On average, xx\% of variation in the output of the frequency channels was explained by the average across channels. An example of such a cross-channel sum is plotted in figure \ref{fig:processing}xx. In correspondence to our previous work \citep{Vanderelst2016}, we refer to this average across frequency channels as a template.

@@

In our previous work, the ensonification device was used to ensonify bat habitats at three different sites. However, here we only use data collected along two linear pathways through natural habitats. At the first site, a public park in Sde Boker, Israel (Fig. \ref{fig:datacollection}). The environment consisted of a walkway lined with rocks and vegetation. Unpublished bat monitoring data at the site confirmed that bat used this corridor to commute. The ensonification device was placed at 50 positions along a straight line spanning a total of 10 meters Locations were spaced 20 cm apart. At each location, echoes were collected for 31 azimuth directions (-150 to +150 degrees) and 7 elevation directions (-40 to +25 degrees). Therefore, at each location, echoes were collected from 31 x 7 = 217 directions. At each azimuth and elevation direction, three repeat measurements were recorded. In this way, a total of 32,550 echo trains (50 positions x 217 directions x 3 repeats) for each of the 31 microphones in the device were collected.

@@

Training a different neural network for each of these features ensured that the training algorithm paid equal attention to each feature. When training a single network for all features, the optimizer could preferably reduce errors in the location predictions as this feature has the largest variance. The full architecture can be considered as a single network with different streams for each feature (fig. \ref{fig:networks}).

We should point out that we did not take any steps to avoid so-called overfitting by the neural networks. Overfitting occurs when a neural network \citep{Ghotra2017}, or other mathematical models \citep{Hawkins2004}, extract complicated relationships between inputs and outputs that do not hold beyond the training examples. Overfitting typically reduces the ability of the model to generalize its predictive power to new data. However, in this case, the model is not required to generalize. Instead, we are only interested in whether the neural networks can learn and classify the set of training examples.

@@

Figure \ref{fig:mechanism} shows the number of \pcs\ needed to capture 99\% of the templates variance. From this plot, it can be seen that the number of \pcs\ increases when shuffling the echo sequences from xx to xx. Omitting the 1 kHz filter from the  \citet{Wiegrebe2008} model increases the number of \pcs\ to xx. Inspecting the average spectrum of the templates (fig. \ref{fig:mechanism}, inset) shows that most energy in the templates is contained in frequencies below 1 kHz. Therefore, if we take 1 kHz as the highest relevant frequency, it ought to be possible to completely represent templates by sampling them at 2 kHz. The corresponds to about 70 samples per template, as templates consist of 7499 samples at 219 kHz sample rate. This corresponds roughly to the number of \pcs\ required to represent templates derived from echo sequences that were shuffled  (fig. \ref{fig:mechanism}). The number of \pcs\ required to represent unshuffled templates is about half of this, i.c., \pca. Therefore, we conclude that the 1 kHz filter is the main, but not exclusive, locus of the dimensionality reduction. The structure of the echo signals also decreases the dimensionality of the templates. Therefore, we conclude that the sonar system of bats, through a number of filtering stages, compresses the 3D complexity of the world onto a lower dimensional representation.

\begin{figure}[tb]
	\centering
	\includegraphics[width=1\linewidth]{../results/mechanism}
	\caption{}
	\label{fig:mechanism}
\end{figure}

Despite the low dimensional character of the templates, they are sufficient to recognize environments. Indeed, this was demonstrated in our previous work \citep{Vanderelst2016}. Here, we directly demonstrated that the \pca\ principal components are sufficient to discriminate their associated locations using neural networks. The neural networks' performance was somewhat that the perfect memory approach we used previously (figs. \ref{fig:israelperformance}, \ref{fig:royalperformance}).

We showed that a low dimensional signal is sufficient for recognizing environments. This is a critical component of navigation. However, bats do more than navigating. They also rely on sonar for flight control. This includes, avoiding obstacles or following guiding structures. The environmental description the bat has access to for flight control has the same low-level character than that used for place recognition. Therefore, our current work suggests that this low-level description should be sufficient for flight control as well. Indeed, we have previously demonstrated that a low-dimensional description could be sufficient for obstacle avoidance and corridor following \citep{Vanderelst2015a,Mansour2019}. In that work, we converted both the echoes at the left and at the right ear to template-like representation. Subsequent work will have to demonstrate that the parameters of the \pca D space presented here can be used directly for controlling flight. In other words, we conjecture it should be possible to extract \pca\ features from an echo signal and convert this vector into a motor command. One can presume a neural network that converts the \pca\ features to flight control parameters.

@@

>>>>

Echolocating bats are long-lived, mobile animals. They use the landscape intensively for foraging, roosting, and commuting. They have excellent spatial memory \citep{Barchi2013,VonHelversen2005} and can rely on sonar to navigate to notable locations like foraging grounds, drinking places \citep[see][for references]{Vanderelst2016,Vanderelst2017}. Furthermore, bats have been reported to use sonar-based recognition of landmarks, both under experimental \citep{Jensen2005,Yu2019}, and natural \citep{Verboom1999} conditions. Echolocating bats can navigate to salient places \citep{Schnitzler2003}, relying solely on their sonar system. Indeed, bats deprived of sight have been found to successfully return to their roost when displaced by up to 40 miles \citep{Stones1969,Williams1966,Davis1970,Mueller1957}.

@@

Together these results suggest that echolocating bats live in a low dimensional perceptual world. The physics of the sonar system and the auditory periphery map the complexity of the 3D environment onto a small number of uncorrelated features, irrespective of whether we use artificial or real echo sequences. Nevertheless, this limited set of features is sufficiently rich for bats to recognize locations, and as we discuss, could also support flight control.

@@

We created artificial echo sequences by randomly spacing ideal point reflectors. The distances of the point reflectors were selected using a two-step process. First we selected $n_{s}$ seed points $s$, uniformly distributed over a range from 0.5 to 7 meters. Next, for every seed point $s$, we selected $n_{c}$ cloud points normally distributed around $s$, with standard deviation 0.25 m. We added a single reflector at a distance between xx and xx to each collection of points. This two step process to modeling natural reflectors was inspired by the results of \citet{Yovel2009}, who found that echoes returning from vegetation could be better modeled using a similar two-step process than simply using uniformly spaced points. We generated xx sets of points for each combination of $n_s \in [xx,xx,xx, xx]$ and $n_c \in [xx, xx, xx, xx]$. Synthetic echoes consisting of few reflectors are likely to consist of only weak echoes.

Each set of points was converted into an environmental impulse response with a sampling frequency $f_s$ of 219 kHz. The environmental impulse response length was set to 7499 samples. These parameters correspond to the recording length and the sampling frequency used in the acoustic data collection (see below). In representing each point reflector by a Dirac impulse, we modeled attenuation of the echoes by introducing an amplitude calculated based on spherical spreading (both ways). The final amplitude of each Dirac impulse was determined by sampling from a normal distribution centered on the calculated amplitude and with a standard deviation of 10 dB. This final operation represents a very simple model of varying reflector strength due e.g., to shape and orientation of the reflecting feature.

The impulse responses were converted to echo sequences by convolving them with a bat-like emission. We used a 1 millisecond long linear frequency-swept cosine starting at 100 kHz and ending at 40 Khz. Again, these parameters correspond to those used in the acoustic data collection described below. The echo sequences were dechirped as follows, 
%
%ifft(complex conj(ftt emission) .* fft(echo))
%
\begin{equation}\label{key}
E_d = \mathcal{F}^{-1}((\mathcal{F}(S))^* . \ \mathcal{F}(E))
\end{equation}
with $\mathcal{F}$ denoting the Fourier transform and $()^*$ the complex conjugate. The symbols $S$, $E$, $E_d$ refer to the emission, the echo sequence and the dechirped echo sequence, respectively. Examples of the resulting echo sequences are shown in figures \ref{fig:processing} and \ref{fig:synthetictemplates}.

Artificial echo sequences were filtered using a model of the bat's auditory periphery, as proposed by \citet{Wiegrebe2008}. This model simulates the temporal and spectral resolution of the bat's cochlea. The model consists of a gammatone filterbank followed by half-wave rectification and non-linear compression with an exponent of 0.4. Finally, the model includes a 1 kHz low-pass filter modeling the response characteristics of the inner hair cells. In the current implementation, we used a gammatone filterbank with xx filters, centered around xx kHz. The filters were spaced to correspond to equivalent rectangular bandwidths \citep{Glasberg1990}. The output of each channel was attenuated with a time (distance) weighted factor corresponding to the atmospheric attenuation at the channel's center frequency. This operation approximated frequency-dependent atmospheric attenuation. The first 6 ms of data were removed. This was done to ensure the data corresponded with the measurement data. As indicated below, we removed the first 6 ms of the measurement  data to avoid including the pickup of the emitted signal.

Figure \ref{fig:processing} depicts a single example of a synthetic echo sequence. It also illustrates the resulting response for each of the xx channels of the \citet{Wiegrebe2008} model. From these responses, it can be seen that they are highly correlated across channels. This is confirmed when analyzing the average cross-channel correlation for all xx synthetic echo sequences (Fig. \ref{fig:processing}xx). We calculated the correlation $\bar{\rho}$ between the output $X_i$ of each frequency channel $i$ and the mean across channels $\bar{X}$ for each synthetic echo, as follows,
%
\begin{equation}
\bar{\rho} = \frac{1}{n} \sum_i \frac{\mbox{cov}(X_i,\bar{X})}{\sigma_{X_i}, \sigma_{\bar{X}}} 
\end{equation}
%
The distribution of $\bar{\rho}$ is plotted in \ref{fig:processing}xx. From this distribution, it can be seen that the value of $\bar{\rho}$ ranged from xx to xx. From these high cross-channel correlations, we conclude that the output of the cochlear model can be summarized by averaging across channels without extensive loss of information.

An example of such a cross-channel sum is plotted in figure \ref{fig:processing}xx. In correspondence to our previous work \citep{Vanderelst2016}, we refer to this average across frequency channels as a template.

@@

Here we use data collected along two linear pathways through natural habitats \citep{Vanderelst2016}. At the first site, a public park in Sde Boker, Israel (Fig. \ref{fig:datacollection}). The environment consisted of a walkway lined with rocks and vegetation. Unpublished bat monitoring data at the site confirmed that bats used this corridor to commute. The ensonification device was placed at 50 positions along a straight line spanning a total of 10 meters Locations were spaced 20 cm apart. At each location, echoes were collected for 31 azimuth directions (-150 to +150 degrees) and 7 elevation directions (-40 to +25 degrees). Therefore, at each location, echoes were collected from 31 x 7 = 217 directions. At each azimuth and elevation direction, three repeat measurements were recorded. In this way, a total of 32,550 echo trains (50 positions x 217 directions x 3 repeats) for each of the 31 microphones in the device were collected.

@@

Training a different neural network for each of these features ensured that the training algorithm paid equal attention to each feature. When training a single network for all features, the optimization procedure would preferably reduce errors in the location predictions as this feature has the largest variance. The full architecture can be considered as a single network with different streams for each feature (fig. \ref{fig:networks}).

We should point out that we did not take any steps to avoid so-called overfitting by the neural networks. Overfitting occurs when a neural network \citep{Ghotra2017}, or other mathematical models \citep{Hawkins2004}, extract complicated relationships between inputs and outputs that do not hold beyond the training examples. Overfitting typically reduces the ability of the model to generalize its predictive power to new data. However, in this case, the model is not required to generalize to other environments. Instead, we are only interested in whether the neural networks can learn and classify the set of training examples. Nevertheless, it would be beneficial if the neural networks would be capable of classifying templates corresponding to directions that are slightly different than those included in the data sets. Indeed, this is a feature of the Euclidean distance measure used in implementing the perfect memory. In that approach, novel templates will be classified to positions with similar templates. We verified whether this properties was also found in the neural networks the case by generating a set of templates by interpolating in azimuth and elevation between the directions ensonified in the data set. In our data sets, we ensonified 31 different azimuth directions and 7 elevation directions. The interpolated data consisted of 30 azimuths and 7 elevations. We used the neural networks trained on the original data to classify these interpolated data. It was found that the neural network indeed classified these interpolated data to azimuths and elevations close to the positions from which they were interpolated (Figure \ref{fig:interpolation}).

@@

Figure \ref{fig:mechanism} shows the number of \pcs\ needed to capture 99\% of the templates variance. From this plot, it can be seen that the number of \pcs\ increases when shuffling the echo sequences from xx to xx. Omitting the 1 kHz filter from the  \citet{Wiegrebe2008} model increases the number of \pcs\ to xx. Including the 1 kHz filter, the average spectrum of the templates (fig. \ref{fig:mechanism}, inset) shows that most energy in the templates is contained in frequencies below 1 kHz. Therefore, if we take 1 kHz as the highest relevant frequency, it ought to be possible to completely represent templates by sampling them at 2 kSamples/sec. This corresponds to about 70 samples per template, as templates consist of 7499 samples at 219 kHz sample rate. This corresponds roughly to the number of \pcs\ required to represent templates derived from echo sequences that were shuffled  (fig. \ref{fig:mechanism}). The number of \pcs\ required to represent unshuffled templates is about half of this, i.c., \pca. Therefore, we conclude that the 1 kHz filter is the main, but not exclusive, locus of the dimensionality reduction. The structure of the echo signals also decreases the dimensionality of the templates. Therefore, we conclude that the sonar system of bats, through a number of filtering stages, compresses the 3D complexity of the world onto a lower dimensional representation.

\begin{figure}[tb]
	\centering
	\includegraphics[width=1\linewidth]{../results/mechanism}
	\caption{Barplot depicting the number of required \pcs\ to capture $>99\%$ of the variance in the templates derived from synthetic echoes. The label \textit{1 kHz filter} refers to whether the 1 kHz filter was omitted in the implementation of the \citet{Wiegrebe2008} model. The label \textit{shuffled} refers to whether the samples of the echoes were randomly shuffled before generating the templates. The inset shows the average spectrum of the templates generated using non-shuffled echoes and a version of the \citet{Wiegrebe2008} model including a 1 kHz filter (i.e., the default model).}
	\label{fig:mechanism}
\end{figure}

Despite the low dimensional character of the templates, they are sufficient to recognize environments. Indeed, this was already demonstrated in our previous work \citep{Vanderelst2016}. Here, we directly demonstrated that the \pca\ principal components are sufficient to discriminate the associated locations and viewing directions using neural networks. The neural networks' performance was somewhat better than the perfect memory approach we used previously (figs. \ref{fig:israelperformance}, \ref{fig:royalperformance}).

While recognizing location in the environment is a critical component of navigation, bats use their sonar not only for navigating. They also rely on sonar for flight control. This includes, avoiding obstacles and/or following guiding structures. The environmental description the bat has access to for flight control starts from the same cochlear output  used for place recognition. Therefore, our current work suggests that the same low-dimensional description should be sufficient for flight control as well. Indeed, we have previously demonstrated that a low-dimensional description is sufficient for obstacle avoidance and corridor following \citep{Vanderelst2015a,Mansour2019}. In that work, we converted both the echoes at the left and at the right ear into a template-like representation used as input to the motion controller. Subsequent work will have to demonstrate whether the parameters of the \pca D space presented here can be used directly for controlling flight. In other words, we conjecture it should be possible to extract \pca\ features from an echo signal and convert this vector into a motor command. One could presume a neural network that converts the \pca\ features to flight control parameters.

@@

\begin{figure}[h]
	\centering
	\includegraphics[width=1\linewidth]{../results/interpolation}
	\caption{sadasasasdasd}
	\label{fig:interpolation}
\end{figure}
